---
title: "Biostatistical Methods II Homework 3"
author: "Arghya Kannadaguli (ak5357)"
date: "2025-03-11"
output: html_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(dplyr)
library(stringr)
```

## **Problem 1**

**(50%) The following table gives data from a retrospective study between esophageal cancer and daily alcohol consumption adjusted for age.**

```{r}
# Create dataframe
cancer_data = tibble(
  age = c(25, 35, 45, 55, 65, 75), #age data modeled as continuous, despite categorical format from table
  case_dac_low = c(0, 5, 21, 34, 36, 8),
  case_dac_high = c(1, 4, 25, 42, 19, 5),
  ctrl_dac_low = c(106, 164, 138, 139, 88, 31),
  ctrl_dac_high = c(9, 26, 29, 27, 18, 0)
)

knitr::kable(cancer_data)
```

**(a) Fit a prospective model to the data to study the relation between alcohol consumption, age, and disease (model age as a continuous variable taking values 25, 35, 45, 55, 65, and 75). Interpret the result.**

First let's pivot the table so it's easier to use.
```{r}
cancer_data = cancer_data |> 
  pivot_longer(
    cols = -age,
    names_to = "temp",
    values_to = "count"
  ) |> 
  mutate(
    status = factor(str_extract(temp, "case|ctrl"), levels = c("ctrl", "case")),
    dac = factor(str_extract(temp, "low|high"), levels = c("low", "high"))
  ) |> 
  select(-temp) |> 
  relocate(-count)

head(cancer_data)
```

If each row signified an individual rather than a group, we would use a Bernoulli model. However, since this is grouped data, we can use a binomial model.















